Certified Symbolic Transducer with the Application of String Solver

### 2025-02-10 Paper/main.tex
polish and proceed the paper

\begin{figure}[t]
	\begin{lstlisting}
record (|$'q,$||$~'a$|) |$\mathbf{NFA}$| =
	|$\mathcal{Q}$| :: "|$'q$| set"
	|$\Delta$| :: "(|$'q,$||$~'a$|) LTS"
	|$\mathcal{I}$| :: "|$'q$| set"
	|$\mathcal{F}$| :: "|$'q$| set"

@@

\begin{figure}[ht]
	\begin{lstlisting}
definition productT :: |$"('q,'a,'b,'c)"$| NFT |$\Rightarrow$| ('q,'a) NFA |$\Rightarrow$| 
           |$(('a,'b)$| Tlabel |$\Rightarrow 'a \text{ set}\Rightarrow 'b\text{ set option})\Rightarrow$|
           |$('q\times 'q, ~'b)$| eNFA where
  "productT |$\mathcal{T}$| |$\mathcal{A}$| F = |$\llparenthesis$|
    |$\mathcal{Q}e = \mathcal{Q}_t~\mathcal{T} \times \mathcal{Q} ~\mathcal{A}$|,
    |$\Delta_e = \{((p,p'),~\texttt{the }(((\mathcal{M}~\mathcal{T})~f)\texttt{ None}),~(q, p'))\mid p, p', q, f.~p'\in \mathcal{Q}~\mathcal{A} \land$|
          |$(p, (\texttt{None}, f), q)\in \Delta_T \mathcal{T} \land \exists S. (\mathcal{M}~\mathcal{T})~f~\text{None} = \text{Some } S\} ~\cup$|
        |$\{((p,p'),~\texttt{the }(F~((\mathcal{M}~\mathcal{T})~f)~(\sigma_1\cap\sigma_2)),~(q, q'))\mid p, p', q, \sigma_1, \sigma_2, q', f.~$|
          |$ (p, (\text{Some }\sigma_1, f), q)\in \Delta_T~\mathcal{T} \land (p', \sigma_2, q')\in\Delta~\mathcal{A}\land \sigma_1\cap \sigma_2 = \emptyset \land$|
            |$ \exists S.~F~((\mathcal{M}~\mathcal{T})~f)~(\sigma_1\cap\sigma_2) = \text{Some } S\}$|,
    |$\Delta_e' = \{((p,p'),~\text{the }(((\mathcal{M}~\mathcal{T})~f)\texttt{ None}),~(q, p'))\mid p, p', q, f.~p'\in \mathcal{Q}~\mathcal{A} \land$|
          |$(p, (None, f), q)\in \Delta_T~\mathcal{T} \land (\mathcal{M}~\mathcal{T})~f~\texttt{None} = \text{None}\} ~\cup$|
        |$\{((p,p'),~\texttt{the }(F~((\mathcal{M}~\mathcal{T})~f)~(\sigma_1\cap\sigma_2)),~(q, q'))\mid p, p', q, \sigma_1, \sigma_2, q', f.~$|
          |$ (p, (\text{Some }\sigma_1, f), q)\in \Delta_T~\mathcal{T} \land (p', \sigma_2, q')\in\Delta~\mathcal{A}\land \sigma_1\cap \sigma_2 = \emptyset \land$|
            |$ \exists x\in(\sigma_1\cap\sigma_2).~((\mathcal{M}~\mathcal{T})~f)~(\text{Some } x) = \text{None} \}$|,
    |$\mathcal{I}_e = \mathcal{I}_t~\mathcal{T} \times \mathcal{I}~\mathcal{A}$|
    |$\mathcal{F}_e = \mathcal{F}_t~\mathcal{T} \times \mathcal{F}~\mathcal{A}$| |$\rrparenthesis$|"
	\end{lstlisting}
\caption{The formalization of $\varepsilon-$NFA and NFA in Isabelle/HOL}
\label{fig-def-FTProd}
\end{figure}

@@

\begin{figure}[ht]
	\begin{lstlisting}
fun outputE :: "|$('a \text{ option} \times 'b \text{ option}) \text{ list} \Rightarrow 'b \text{ list}$|" where
  "outputE [] = []" |$\mid$|
  "outputE ((_, Some a) # l) = a # (outputE l)" |$\mid$|
  "outputE ((_, None) # l) = (outputE l)"

@@

\begin{figure}[ht]
	\begin{lstlisting}
lemma productT_correct:
  fixes |$\mathcal{T}$| |$\mathcal{A}$| F
  assumes F_ok1: "|$\forall~f~s.~(\forall e \in s. f (\text{ Some } e) = \text{None}) \mapsto \text{ F}~f~s = \text{None}"$|
      and F_ok2: "|$\forall~f~s.~\text{F }f~s = $|
            |$\text{Some }(\cup \{\text{the }(f (\text{ Some }e))\mid e. e \in s \land f~(\text{Some }e) \neq \text{ None}\})$|"
      and wfTA: "NFT_wf |$\mathcal{T}~\land \text{ NFA } \mathcal{A}$|"
    shows "|$\mathcal{L}_e$| (productT |$\mathcal{T}~\mathcal{A}$| F) = outputL |$\mathcal{T}~(\mathcal{M}~\mathcal{T})~\mathcal{A}$|"

@@

\subsection{Intervals}
An interval is defined as a pair $[i, j]$, which corresponds to the set $\{e \mid e \geq i \land e \leq j\}$. In our formalization, we extends the interval definition to a list, that is $[(i_1, j_1), \ldots, (i_n, j_n)]$, which corresponds to the set $\bigcup_{1\leq k\leq n}\{e \mid i_k \leq e \leq i_n\}$. This definition enables us to merge more transitions in NFAs. In the sequent, intervals always mean interval lists.
There are a collection interfaces for intervals. 
\begin{enumerate}
  \item $\texttt{semI}~i$, formalize the semantics of an interval as a set.
  \item $\texttt{empty\_intervals}~i$, checks the emptiness of an interval.
  \item $\texttt{intersectIs}~i_1~i_2$, computes the intersection of two intervals.
  \item $\texttt{diffIs}~i_1~i_2$, computes the intervals difference, that is the interval corresponds to the set $\texttt{semI}(i_1)-\texttt{semI}(i_2)$.
\end{enumerate}

@@

Fig. \ref{fig-compute-nft-product} illustrates the computation of NFT products. \texttt{prods\_imp} computes the product of two sets. It is illustrated in \ref{fig-def-prods_imp}. We use the constructs from \emph{Refine\_Monadic}. The algorithm itself is intuitive: the nested two loops traverse all possible combination of states in the states.

\begin{figure}[ht]
	\begin{lstlisting}
definition productT_impl where
  "productT_impl |$\mathcal{T}$| |$\mathcal{A}$| F fe = do {
    Q |$\leftarrow$| prods_imp (nft_states |$\mathcal{T}$|) (nfa_states |$\mathcal{A}$|);
    (D1, D2) |$\leftarrow$| trans_comp_imp (nft_tranfun |$\mathcal{T}$|) F fe 
        (nft_trans |$\mathcal{T}$|) (nfa_trans |$\mathcal{A}$|) (nfa_states |$\mathcal{A}$|);
    I |$\leftarrow$| prods_imp (nft_initial |$\mathcal{T}$|) (nfa_initial |$\mathcal{A}$|);
    F |$\leftarrow$| prods_imp (nft_accepting |$\mathcal{T}$|) (nfa_accepting |$\mathcal{A}$|);
    RETURN (Q,D1,D2,I, F)
  }"
\end{lstlisting}
\caption{The computation of NFT product}
\label{fig-compute-nft-product}
\end{figure}

\begin{figure}[ht]
	\begin{lstlisting}
definition prods_imp where
  "prods_imp Q1 Q2 =
   FOREACH {q. q |$\in$| s.|$\alpha$| Q1} (|$\lambda$| q Q. do {
    S |$\leftarrow$| FOREACH {q. q |$\in$| s.|$\alpha$| Q2}
         (|$\lambda$| q' Q'. RETURN (ss.ins (q,q') Q')) (ss.empty ());
    RETURN (ss.union Q S)
   }) (ss.empty ())"
\end{lstlisting}
\caption{The formalization of $\varepsilon-$NFA and NFA in Isabelle/HOL}
\label{fig-def-prods_imp}
\end{figure}

The challenging part is how to compute the transitions sets. It is illustrated in Fig. \ref{fig-def-prods-imp}. The function \texttt{trans\_comp\_imp} traverse the transtions in the NFT and split each transtion into two cases. This function uses another two functions:

1. \texttt{copy\_tran\_impl}

2. \texttt{copy\_tran'\_impl}

\begin{figure}[ht]
	\begin{lstlisting}
definition trans_comp_imp where
  "trans_comp_imp M F fe T1 T2 Q =
    FOREACH {t. t |$\in$| dt.|$\alpha$| T1}
     (|$\lambda$| (q, (|$\alpha$|, f), q') (D1, D2). 
       (if (|$\alpha$| = None) then 
          (if ((M f) None = None) then 
            do {
              D2' |$\leftarrow$| copy_tran_imp q q' Q;
              RETURN (D1, ssd.union D2 D2')
            }
            else do {
              D1' |$\leftarrow$| copy_tran'_imp q (the (M f None)) q' Q;
              RETURN (ddt.union D1 D1', D2)
            })
          else 
            (subtrans_comp_imp M q (the |$\alpha$|) f q' F fe T2 D1 D2)
        )) (ddt.empty, ssd.empty ())"

\end{lstlisting}
\caption{The computation of transition sets}
\label{fig-def-prods-imp}
\end{figure}

\section{Refinement to Efficient Data Structure}

@@

@@

The semantics of the replacement operation can be described in two cases: (1) if there are strings in \texttt{str} that match \texttt{pattern}, then replace them with \texttt{replacement}. (2) If there are no matches, then the result is \texttt{str}.

The first case can be modeled by the SFTs. We illustrate this case by an example. Given a replacement operation $\texttt{replace}(s, $"$[0-9]+$"$, $"$\text{ITP}$"$)$, we can model this operation by an SFT.

@@

\begin{enumerate}
  \item Generate an SNFA from the regular expression "$[0-9]+$". Adding output functions to the transitions of the SNFA that outputs empty characters to make it an SFT.
  \item Generate an SNFA from the replacement string "$"$\text{ITP}$"$". Note that a single string is also a regular expression. Adding input labels and output functions to the transitions of the SNFA to make it an SFT.
  \item Concatenate the two SFTs and adding preceding and succeeding transitions to make it accept the whole input string.
\end{enumerate}

We explain it step by step.

\noindent\emph{Step 1.}
The SNFA corresponding to the regular expression "$[0-9]+$" is shown in Figure \ref{fig-snfa-pattern}. On the left side is the SNFA and on the right side is the corresponding SFT. The function \texttt{f} is the output function that outputs empty characters. It can be defined as $\texttt{f} = \lambda x.~\texttt{None}$. The function \texttt{g} is the output function that outputs the characters in the replacement string. It can be defined as (the ascii code of $'I'$, $'T'$, $'P'$ are 73, 84, 80 respectively):

\[
\texttt{g} = \lambda~i~x.~\texttt{match }~i~ with~ 1 \rightarrow [(73, 73)] ~|~ 2 \rightarrow [(84, 84)] ~|~ 3 \rightarrow [(80, 80)] ~|~ \_ ~\rightarrow \texttt{None}
\]

\begin{figure}[h] \centering
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, auto]
  \node[state, initial] (q0)   {$q_0$}; 
  \node[state, accepting] (q1) [right=of q0] {$q_1$};

@@

\noindent\emph{Step 2.}
The SNFA corresponding to the replacement string "\text{ITP}" is shown in Figure \ref{fig-snfa-replacement}. On the left side is the SNFA and on the right side is the corresponding SFT. The function \texttt{f} is the output function that outputs the characters in the replacement string. It can be defined as $\texttt{f} = \lambda x.~\texttt{Some } x$.

\begin{figure}[h] \centering
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, auto]
  % Original NFA for "ITP"
  \node[state, initial] (q0)   {$q_0$}; 
  \node[state] (q1) [right=of q0] {$q_1$}; 
  \node[state] (q2) [below=of q0] {$q_2$}; 
  \node[state, accepting] (q3) [right=of q2] {$q_3$};

@@

  \path[->] 
  (q0') edge node {I/\texttt{g 1}} (q1')
  (q1') edge node {T/\texttt{g 2}} (q2')
  (q2') edge node {P/\texttt{g 3}} (q3');
\end{tikzpicture}
\caption{The SFA for "$"\text{ITP}"$" duplicated}
\label{fig-snfa-replacement}
\end{figure}

\noindent\emph{Step 3.}
The concatenation of the two SFTs is shown in Figure \ref{fig-rearranged-automata}. It contains two steps: (1) rearrange the transitions of the two SFTs to make them accept the whole input string. (2) add the preceding and succeeding transitions to make it accept the whole input string.

\begin{figure}[h] \centering
  \begin{tikzpicture}[shorten >=1pt, node distance=2cm, auto]
    % Second automaton from the first figure
    \node[state, initial] (q0') {$q_0'$}; 
    \node[state] (q1') [right=of q0'] {$q_1'$};

     \path[->] 
     (q0') edge [bend left] node {[0-9]/\texttt{f}} (q1')
     (q1') edge [loop above] node {[0-9]/\texttt{f}} (q1')
     (q0') edge [loop below] node {[\texttt{None}/\texttt{f}]} (q0');

@@

    \path[->] 
    (q0'') edge node {I/\texttt{g 1}} (q1'')
    (q1'') edge node {T/\texttt{g 2}} (q2'')
    (q2'') edge node {P/\texttt{g 3}} (q3'')
    (q3'') edge [loop below] node {[\texttt{None}/\texttt{f}]} (q3'')
    (q1') edge node {[\texttt{None}/\texttt{f}]} (q0'');
  \end{tikzpicture}
  \caption{The SFT for the replacement operation $\texttt{replace}(s, $"$[0-9]+$"$, $"$\text{ITP}$"$)$}
  \label{fig-rearranged-automata}
  \end{figure}

>>>>

\begin{figure}[hbt!]
	\begin{lstlisting}
record (|$'q,$||$~'a$|) |$\mathbf{NFA}$| =
	|$\mathcal{Q}$| :: "|$'q$| set"
	|$\Delta$| :: "(|$'q,$||$~'a$|) LTS"
	|$\mathcal{I}$| :: "|$'q$| set"
	|$\mathcal{F}$| :: "|$'q$| set"

@@

\begin{figure}[hbt!]
	\begin{lstlisting}
definition productT :: |$"('q,'a,'b,'c)"$| NFT |$\Rightarrow$| ('q,'a) NFA |$\Rightarrow$| 
           |$(('a,'b)$| Tlabel |$\Rightarrow 'a \text{ set}\Rightarrow 'b\text{ set option})\Rightarrow$|
           |$('q\times 'q, ~'b)$| eNFA where
  "productT |$\mathcal{T}$| |$\mathcal{A}$| F = |$\llparenthesis$|
    |$\mathcal{Q}e = \mathcal{Q}_t~\mathcal{T} \times \mathcal{Q} ~\mathcal{A}$|,
    |$\Delta_e = \{((p,p'),~\texttt{the }(((\mathcal{M}~\mathcal{T})~f)\texttt{ None}),~(q, p'))\mid p, p', q, f.~p'\in \mathcal{Q}~\mathcal{A} \land$|
          |$(p, (\texttt{None}, f), q)\in \Delta_T \mathcal{T} \land \exists S. (\mathcal{M}~\mathcal{T})~f~\text{None} = \text{Some } S\} ~\cup$|
        |$\{((p,p'),~\texttt{the }(F~((\mathcal{M}~\mathcal{T})~f)~(\sigma_1\cap\sigma_2)),~(q, q'))\mid p, p', q, \sigma_1, \sigma_2, q', f.~$|
          |$ (p, (\text{Some }\sigma_1, f), q)\in \Delta_T~\mathcal{T} \land (p', \sigma_2, q')\in\Delta~\mathcal{A}\land \sigma_1\cap \sigma_2 = \emptyset \land$|
            |$ \exists S.~F~((\mathcal{M}~\mathcal{T})~f)~(\sigma_1\cap\sigma_2) = \text{Some } S\}$|,
    |$\Delta_e' = \{((p,p'),~\text{the }(((\mathcal{M}~\mathcal{T})~f)\texttt{ None}),~(q, p'))\mid p, p', q, f.~p'\in \mathcal{Q}~\mathcal{A} \land$|
          |$(p, (None, f), q)\in \Delta_T~\mathcal{T} \land (\mathcal{M}~\mathcal{T})~f~\texttt{None} = \text{None}\} ~\cup$|
        |$\{((p,p'),~\texttt{the }(F~((\mathcal{M}~\mathcal{T})~f)~(\sigma_1\cap\sigma_2)),~(q, q'))\mid p, p', q, \sigma_1, \sigma_2, q', f.~$|
          |$ (p, (\text{Some }\sigma_1, f), q)\in \Delta_T~\mathcal{T} \land (p', \sigma_2, q')\in\Delta~\mathcal{A}\land \sigma_1\cap \sigma_2 = \emptyset \land$|
            |$ \exists x\in(\sigma_1\cap\sigma_2).~((\mathcal{M}~\mathcal{T})~f)~(\text{Some } x) = \text{None} \}$|,
    |$\mathcal{I}_e = \mathcal{I}_t~\mathcal{T} \times \mathcal{I}~\mathcal{A}$|
    |$\mathcal{F}_e = \mathcal{F}_t~\mathcal{T} \times \mathcal{F}~\mathcal{A}$| |$\rrparenthesis$|"
	\end{lstlisting}
\caption{The formalization of $\varepsilon-$NFA and NFA in Isabelle/HOL}
\label{fig-def-FTProd}
\end{figure}

@@

\begin{figure}[hbt!]
	\begin{lstlisting}
fun outputE :: "|$('a \text{ option} \times 'b \text{ option}) \text{ list} \Rightarrow 'b \text{ list}$|" where
  "outputE [] = []" |$\mid$|
  "outputE ((_, Some a) # l) = a # (outputE l)" |$\mid$|
  "outputE ((_, None) # l) = (outputE l)"

@@

\begin{figure}[hbt!]
	\begin{lstlisting}
lemma productT_correct:
  fixes |$\mathcal{T}$| |$\mathcal{A}$| F
  assumes F_ok1: "|$\forall~f~s.~(\forall e \in s. f (\text{ Some } e) = \text{None}) \mapsto \text{ F}~f~s = \text{None}"$|
      and F_ok2: "|$\forall~f~s.~\text{F }f~s = $|
            |$\text{Some }(\cup \{\text{the }(f (\text{ Some }e))\mid e. e \in s \land f~(\text{Some }e) \neq \text{ None}\})$|"
      and wfTA: "NFT_wf |$\mathcal{T}~\land \text{ NFA } \mathcal{A}$|"
    shows "|$\mathcal{L}_e$| (productT |$\mathcal{T}~\mathcal{A}$| F) = outputL |$\mathcal{T}~(\mathcal{M}~\mathcal{T})~\mathcal{A}$|"

@@

\subsection{Intervals}
An interval is defined as a pair $(i, j)$, which corresponds to the set $\{e \mid e \geq i \land e \leq j\}$. In our formalization, we extends the interval definition to a list, that is $[(i_1, j_1), \ldots, (i_n, j_n)]$, which corresponds to the set $\bigcup_{1\leq k\leq n}\{e \mid i_k \leq e \leq i_n\}$. This definition enables us to merge more transitions in NFAs. Moreover, when an interval operation produces more than two intervals, we do not need to split the intervals. For instance, the interval result of $(1, 5) - (3, 4)$ (the difference between the interval $[1, 5]$ and $[3, 4]$) is equivalent to $[(1, 2), (5, 5)]$.

In the sequent, intervals always mean interval lists.
There are a collection interfaces for intervals. 
\begin{enumerate}
  \item $\texttt{semI}~i$, formalize the semantics of an interval as a set, that is $\texttt{semI }(i,j) = \{e \mid e \geq i \land e \leq j\}$.
  \item $\texttt{emptyI}~i$, checks the emptiness of an interval.
   \item $\texttt{nemptyI}~i$, checks the non-emptiness of an interval.
  \item $\texttt{intersectIs}~i_1~i_2$, computes the intersection of two intervals.
  \item $\texttt{diffIs}~i_1~i_2$, computes the intervals difference, that is the interval corresponds to the set $\texttt{semI}(i_1)-\texttt{semI}(i_2)$.
\end{enumerate}

We define a canonical form for intervals to facilitate the proof. An interval $[i_1, j_1), \ldots, (i_n, j_n)]$ is in canonical form if 
(1) $i_k \leq j_k, 1 \leq k \leq n$,
(2) $j_k\leq i_{k+1}, 1 \leq k < n$.
All the interval operations are proved to keep the canonical form after the operation if the input intervals are in canonical form.

@@

Fig. \ref{fig-compute-nft-product} illustrates the product computation of a SFT and an SFA. \footnote{To make it easier to understand, the Isabelle code presented in this section is not exactly the same as the real one.}
\texttt{prods\_imp} computes the product of two sets. It is illustrated in \ref{fig-def-prods_imp}. We use the constructs from \emph{Refine\_Monadic}. \texttt{prods\_imp} itself is intuitive: the nested two loops traverse all possible combination of states in the states. 
The \texttt{FOREACH} construct in \emph{Refine\_Monadic} is similar to OCaml's \texttt{Set.fold}. $\texttt{FOREACH}~S~f~I$ traverses all elements in $S$ and applies $f$ of type $'a \Rightarrow 'b \Rightarrow 'b$ to each element. $I$ is the initial value of type $'b$.

\begin{figure}[hbt!]
	\begin{lstlisting}
definition productT_impl where
  "productT_impl |$\mathcal{T}$| |$\mathcal{A}$| F fe = do {
    Q |$\leftarrow$| prods_imp (nft_states |$\mathcal{T}$|) (nfa_states |$\mathcal{A}$|);
    (D1, D2) |$\leftarrow$| trans_comp_imp (nft_tranfun |$\mathcal{T}$|) F fe 
        (nft_trans |$\mathcal{T}$|) (nfa_trans |$\mathcal{A}$|) (nfa_states |$\mathcal{A}$|);
    I |$\leftarrow$| prods_imp (nft_initial |$\mathcal{T}$|) (nfa_initial |$\mathcal{A}$|);
    F |$\leftarrow$| prods_imp (nft_accepting |$\mathcal{T}$|) (nfa_accepting |$\mathcal{A}$|);
    RETURN (Q,D1,D2,I, F)
  }"
\end{lstlisting}
\caption{The computation of NFT product}
\label{fig-compute-nft-product}
\end{figure}

\begin{figure}[hbt!]
	\begin{lstlisting}
definition prods_imp where
  "prods_imp Q1 Q2 =
   FOREACH {q. q |$\in$| Q1} (|$\lambda$| q Q. do {
    S |$\leftarrow$| FOREACH {q. q |$\in$| Q2}
         (|$\lambda$| q|$'$| Q|$'$|. RETURN ({(q,q|$'$|)} |$\cup$| Q|$'$|)) |$\emptyset$|;
    RETURN (Q |$\cup$| S)
   }) |$\emptyset$|"
\end{lstlisting}
\caption{The formalization of $\varepsilon-$NFA and NFA in Isabelle/HOL}
\label{fig-def-prods_imp}
\end{figure}

\begin{figure}[hbt!]
	\begin{lstlisting}
definition trans_comp_imp where
  "trans_comp_imp M F T1 T2 Q =
    FOREACH {t. t |$\in$| T1}
      (|$\lambda$| (q, (|$\alpha$|, f), q|$'$|) (D1, D2). 
        (if (|$\alpha$| = None) then 
          (subtrans_comp_|$\varepsilon$| M q  f q|$'$| F T2 D1 D2)
        else 
          (subtrans_comp M q (the |$\alpha$|) f q|$'$| F T2 D1 D2)
      )) (|$\emptyset$|, |$\emptyset$|)"
\end{lstlisting}
\caption{The computation of transition sets}
\label{fig-def-prods-imp}
\end{figure}

The challenging part is how to compute the transition sets. It is illustrated in Fig. \ref{fig-def-prods-imp}. The function \texttt{trans\_comp\_imp} traverses the transitions in the SFT and it calls two the following two functions to compute the synchronized transitions:
(1) \texttt{subtrans\_comp\_$\varepsilon$}: handles the case that the transition in the SFT is $\varepsilon$-transition.
(2) \texttt{subtrans\_comp}: handles the case that the transition in the SFT is not $\varepsilon$-transition.

We elaborate only on the function \texttt{subtrans\_comp} in Fig. \ref{fig-def-subtrans_comp}. The other function is similar. 
This function traverses the transitions SFA's transitions $\texttt{T}$. If it find the a transtion $(q1, \alpha', q1')\in \texttt{T}$ satisfies the following condition:
$\alpha \cap \alpha' \neq \emptyset$ (Checked by \texttt{nemptyIs}),
it started to compute the non-$\varepsilon$-transitions set $\texttt{D1}$ and the $\varepsilon$-transitions set $\texttt{D2}$ by checking whether the output set of $\alpha \cap \alpha'$ is empty or not.

\begin{figure}[hbt!]
	\begin{lstlisting}
definition subtrans_comp_imp where
  "subtrans_comp_imp M q |$\alpha$| f q|$'$| F fe T D1 D2 =
    FOREACH
      {t. t |$\in$| T} (|$\lambda$| (q1, |$\alpha'$|, q1|$'$|) (D1, D2).
        (if (nemptyIs (intersectIs |$\alpha$| |$\alpha'$|)) then do {
          D1 |$\leftarrow$| (if (F (M f) (intersectIs |$\alpha$| |$\alpha'$|)) |$\neq$| None) 
            then RETURN 
            let |$\alpha_i$| = (the (F (M f) (intersectIs |$\alpha$| |$\alpha'$|))) in
            ({(q,q1) |$\alpha_i$| (q|$'$|,q1|$'$|)} |$\cup$| D1)
            else RETURN D1);
          D2 |$\leftarrow$| (if (M f) (intersectIs |$\alpha$| |$\alpha'$|) |$\neq$| |$\emptyset$| then 
            RETURN (((q,q1), (q|$'$|,q1|$'$|)) |$\cup$| D2) else RETURN D2);
          RETURN (D1, D2)
        }
        else (RETURN (D1, D2)))) (D1, D2)"
    \end{lstlisting}
    \caption{The computation of transition sets}
    \label{fig-def-subtrans_comp}
    \end{figure}

  \begin{figure}[hbt!]
    \begin{lstlisting}
  lemma productT_imp_correct:
  assumes finite_TT: "|$\text{finite } (\Delta_T~\mathcal{T})$|"    
      and finite_TA: "|$\text{finite } (\Delta~\mathcal{A})$|"
      and finite_Q: "|$\text{finite } (\mathcal{Q}~\mathcal{A})$|"
      and finite_TQ: "|$\text{finite } (\mathcal{Q}_T~\mathcal{T})$|"
      and finite_I: "|$\text{finite } (\mathcal{I}~\mathcal{A})$|"
      and finite_TI: "|$\text{finite } (\mathcal{I}_T~\mathcal{T})$|"
      and finite_F: "|$\text{finite } (\mathcal{F}~\mathcal{A})$|"
      and finite_TF: "|$\text{finite } (\mathcal{F}_T~\mathcal{T})$|"
  shows "|$\text{productT\_imp } \mathcal{T}~\mathcal{A}~F~\text{fe} \leq \text{SPEC } (\lambda A.~A = \text{productT } \mathcal{T}~\mathcal{A}~F)$|"
  \end{lstlisting}
  \caption{The refinement of the product computation}
  \label{fig-def-productT_imp_correct}
  \end{figure}

  To prove the correctness of the product computation introduced above, we just need to show that the product result of \texttt{productT\_imp} in Fig. \ref{fig-def-prods_imp} is a refinement of the product result of \texttt{productT} in Figure \ref{fig-compute-nft-product}. This refinement specification is illustrated in Figure \ref{fig-def-productT_imp_correct}. The proof is under the framework of \emph{Refine\_Monadic}. The specification $\mathbf{C} \leq\texttt{SPEC}~\mathbf{A}$ means that $\mathbf{C}$ is an element in the set $\mathbf{A}$. In Fig. \ref{fig-def-productT_imp_correct}, "$\lambda A.~A = \text{productT } \mathcal{T}~\mathcal{A}~F$" specifies a set of elements $A$, such that $A$ equals to $\texttt{productT}~\mathcal{T}~\mathcal{A}~F$. 
  The specification will force us to prove the result $\varepsilon$-NFAs of \texttt{productT\_imp}  and \texttt{productT} have the same sets of initial states, transition relations, $\varepsilon$-transitions, and accepting states.

\subsection{Refinement to Efficient Data Structure}

The last subsection shows the way to compute the product of two NFAs and the transition labels have been refined to interval algebra. But it still uses the data structure of sets in Isabelle/HOL.

The \emph{Refine\_Monadic} framework provides a way to refine the data structure of sets to more efficient data structure, such as red-black trees. The basic idea is to provide a set of \emph{set} operations on the refined data structure and prove the refinement relation between the operations on the refined data structure and the operations on the abstract data structure.
%
In our formalization, we refine the data structure of sets to red-black trees.

@@

CertiStr is an existing string solver based on SFAs. However, due to the limitation of SFAs, it cannot solve the string constraints with replacement operations. CertiStr is based on forward-propagations. We will not elaborate and repeat the idea of forward-propagation in this paper. In this section, we only focus on how to model the replacement operation in SFTs.

\subsection{Modeling the Replacement Operation}

@@

The semantics of the replacement operation can be described in two cases: (1) if there are a substring in \texttt{str} that match \texttt{pattern}, then replace it with \texttt{replacement}. (2) If there are no matches, then the result is \texttt{str} itself.

The first case can be modeled by SFTs. We illustrate this case by an example. Given a replacement operation $\texttt{replace}(s, $"$[0-9]+$"$, $"$\text{ITP}$"$)$, we can model this operation by an SFT.

@@

\begin{enumerate}
  \item Generate an SFA from the regular expression "$[0-9]+$". Adding output functions to the transitions of the SFA that output empty symbol (\texttt{None}) to make it an SFT.
  \item Generate an SNFA from the replacement string "$"$\text{ITP}$"$". Note that a single string is also a regular expression. Add input labels and output functions to the transitions of the SNFA to make it an SFT.
  \item Concatenate the two SFTs and add preceding and succeeding self-loop transitions to make it accept the whole input string.
\end{enumerate}

We explain it step by step by the example.

\noindent\emph{Step 1.}
The SNFA corresponding to the regular expression "$[0-9]+$" is shown in Figure \ref{fig-snfa-pattern}. On the left side is the SNFA and on the right side is the corresponding SFT. The function \texttt{f} is the output function that outputs empty symbol. It can be defined as $\texttt{f} = \lambda x.~\texttt{None}$.

\begin{figure}[h] \centering
\begin{tikzpicture}[shorten >=1pt, node distance=1.5cm, auto]
  \node[state, initial] (q0)   {$q_0$}; 
  \node[state, accepting] (q1) [right=of q0] {$q_1$};

@@

\noindent\emph{Step 2.}
The SNFA corresponding to the replacement string "\text{ITP}" is shown in Figure \ref{fig-snfa-replacement}. On the left side is the SFA and on the right side is the corresponding SFT. The function \texttt{g} is the output function that outputs the characters in the replacement string. It can be defined as (the ascii code of $'I'$, $'T'$, $'P'$ are 73, 84, 80 respectively):
\[
\texttt{g} = \lambda~i~x.~\texttt{match }~i~ with~ 1 \rightarrow [(73, 73)] ~|~ 2 \rightarrow [(84, 84)] ~|~ 3 \rightarrow [(80, 80)] ~|~ \_ ~\rightarrow \texttt{None}
\]

\begin{figure}[h] \centering
\begin{tikzpicture}[shorten >=1pt, node distance=1.5cm, auto]
  % Original NFA for "ITP"
  \node[state, initial] (q0)   {$q_0$}; 
  \node[state] (q1) [right=of q0] {$q_1$}; 
  \node[state] (q2) [below=of q0] {$q_2$}; 
  \node[state, accepting] (q3) [right=of q2] {$q_3$};

@@

  \path[->] 
  (q0') edge node {\texttt{None}/\texttt{g 1}} (q1')
  (q1') edge node {\texttt{None}/\texttt{g 2}} (q2')
  (q2') edge node {\texttt{None}/\texttt{g 3}} (q3');
\end{tikzpicture}
\caption{The SFA for "$"\text{ITP}"$" duplicated}
\label{fig-snfa-replacement}
\end{figure}

\noindent\emph{Step 3.}
The concatenation of the two SFTs is shown in Figure \ref{fig-rearranged-automata}. It contains two minor steps: (1) add a transtion (\texttt{None}/\texttt{f}) from each accepting state of the first SFT to the initial state of the second SFT. (2) add the preceding and succeeding transitions to make it accept the whole input string.

\begin{figure}[h] \centering
  \begin{tikzpicture}[shorten >=1pt, node distance=1.5cm, auto]
    % Second automaton from the first figure
    \node[state, initial] (q0') {$q_0'$}; 
    \node[state] (q1') [right=of q0'] {$q_1'$};

     \path[->] 
     (q0') edge [bend left] node {[0-9]/\texttt{f}} (q1')
     (q1') edge [loop above] node {[0-9]/\texttt{f}} (q1')
     (q0') edge [loop below] node {\texttt{None}/\texttt{f}} (q0');

@@

    \path[->] 
    (q0'') edge node {\texttt{None}/\texttt{g 1}} (q1'')
    (q1'') edge node {\texttt{None}/\texttt{g 2}} (q2'')
    (q2'') edge node {\texttt{None}/\texttt{g 3}} (q3'')
    (q3'') edge [loop below] node {\texttt{None}/\texttt{f}} (q3'')
    (q1') edge node {\texttt{None}/\texttt{f}} (q0'');
  \end{tikzpicture}
  \caption{The SFT for the replacement operation $\texttt{replace}(s, $"$[0-9]+$"$, $"$\text{ITP}$"$)$}
  \label{fig-rearranged-automata}
  \end{figure}

  With this SFT, we can compute the forward image of the replacement operation.
  Let $s' = \texttt{replace}(s, $"$[0-9]+$"$, $"$\text{ITP}$"$)$ be a string constraint and we want to compute the domain of $s'$ regarding this replacement operation.
  Assume the SFT generated from this replacement operation is $\mathcal{T}$ and the string $s$'s domain ranges over an SFA $\mathcal{A}$. The foward image of this replacement operation is $\mathcal{T}~\mathcal{A}$. Therefore, the domain of $s'$ is $\mathcal{T}~\mathcal{A}$.

\subsection{Experiments}

We have implemented a prototype of the string solver to support the replacement operation based on CertiStr[1], which is called CertiStr-R in this paper.

CertiStr-R supports two kinds of replacement operations: (1) \texttt{str.replace s p r}, where the pattern \texttt{p} is a string and (2) \texttt{str.replace\_re s p r}, where the pattern \texttt{p} is a regular expression.

The limitation of CertiStr-R is that it does not support \texttt{replace\_all} and \texttt{replace\_re\_all}. CertiStr-R is possible to extend in the future. Another limitation is that the semantics of \texttt{replace} and \texttt{replace\_re} are different from the definition in SMTLIB.

\section{Related Work}

\section{Conclusion}
